Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-06 tex/main.tex
related work

The translation modifies $\varphi$ in two main ways. First, it restricts the
quantification on function definitions for $\con{f}$ to the corresponding
uninterpreted type $\fargsort{\con{f}}$ and inserts appropriate coercions
$\fargx{\con{f}}{i}$. Second, it augments $\varphi$ with additional constraints
$\exists a : \fargsort{\con{f}}.\;\, \vecfarg{\con{f}}( a ) \teq \vec t$; these
ensure that the restricted definition for $\con{f}$ covers all relevant tuples
of terms, namely the ones for which an $\con{f}$ application exists and is
relevant to the satisfiability of $\varphi$.
%\jb{what does ``current branch of'' mean? can we drop it / rephrase it / vague
%it up?}

@@

\begin{verbatim}
first-order model finders: SEM, MACE/Mace, Paradox, FMDarwin, iProver
  either direct search or reduction to SAT

Alloy (and Alloy idiom by Kuncak \& Jackson), Kodkod

The model finder for KIV based on Alloy

Refute
  * unsound handling of datatypes

Nitpick
  * nice: three-valued logic (e.g. local overflow)
    * solving a different problem:
      * no built-in notion, what's approximated is an idealized notion, no
        injections $\concret$
  * not so nice:
    one abstract domain per type, not per function argument
  * guards + one-domain per arg are a good combi, guards + one-single-domain is bad because of
    odd schemes, give unsat example
    [* show that it works also for example with weird recursion schemes]

  * other features:
    * inductive / coinductive predicates
    * quotient types

  * abstract interpretation

Leon

in smt : z3, cvc4 (briefly, text in main section gives detail)
\end{verbatim}

@@

We presented a translation scheme that makes more SMT problems amenable to
finite model findings in SMT. In conjunction with the finite model finding
techniques introduced by Reynolds et al.\
\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013}, it makes it possible to
find models of quantified formulas over infinite types, such as arithmetic
types and recursive datatypes.

in definitional form with respect to admissible function definitions

Universal quantifiers ranging over infinite types

\begin{verbatim}
  * future work
    * more eval
      * against other non-SMT-solvers (notably Leon and Nitpick)
      * problems exercising codatatypes (which \cvc supports \cite{xxx}) and
        corecursive functions
    * more features
      * e.g. encode (co)inductive predicates as (co)recursive functions
\end{verbatim}

>>>>

The translation modifies $\varphi$ in two main ways. First, it restricts the
quantification on function definitions for $\con{f}$ to the corresponding
uninterpreted type $\fargsort{\con{f}}$ and inserts appropriate coercions
$\fargx{\con{f}}{i}$. Second, it augments $\varphi$ with additional exisential
constraints
$\exists a : \fargsort{\con{f}}.\;\, \vecfarg{\con{f}}( a ) \teq \vec t$; these
ensure that the restricted definition for $\con{f}$ covers all relevant tuples
of terms, namely the ones for which an $\con{f}$ application exists and is
relevant to the satisfiability of $\varphi$.
%\jb{what does ``current branch of'' mean? can we drop it / rephrase it / vague
%it up?}

@@

We described the most closely related work, by Ge and de Moura
\cite{GeDeM-CAV-09} and by Reynolds et al.\
\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013}, in the introduction already.
Outside the world of SMT, model finding and counterexample generation have been
studied many times. The text below is partly adapted from Blanchette's Ph.D.\
thesis \cite{blanchette-2012-phd}.

Some finite model finders are based on a reduction to Boolean satisfiability
(SAT). They translate the input problem to propositional logic and pass it to a
SAT solver. The translation is parameterized by upper or exact finite bounds on
the cardinalities of the atomic types. This procedure was pioneered by McCune
in the earlier versions of Mace (also capitalized MACE) \cite{mccune-1994}.
Other SAT-based finders are Paradox \cite{claessen-sorensson-2003}, FMDarwin
\cite{xxxx}, and iProver \cite{xxxx} for first-order logic with equality; the
Alloy Analyzer and its back-end Kodkod \cite{torlak-jackson-2007} for
first-order relational logic; and Refute \cite{weber-2008} and Nitpick
\cite{blanchette-nipkow-2010} for higher-order logic.

An alternative to translating the problem to propositional logic is to perform
an exhaustive model search directly on the original problem. Given fixed
cardinalities, the search space is represented in memory as multidimensional
arrays. The procedure tries different values in the function and predicate
table entries, checking each time if the problem is satisfied and backtracking
if necessary. This approach was pioneered by SEM \cite{zhang-zhang-1995} and
FINDER \cite{slaney-1994} and serves as the basis of many more model finders,
notably the Alloy Analyzer's precursor \cite{jackson-1996} and the later
versions of~Mace \cite{mccune-prover9-mace4}.

Most of the above tool cannot cope with recursive datatypes or other infinite
types. Kuncak and Jackson \cite{kuncak-and-jackson-2005} presented an idiom for
encoding datatypes and recursive functions in Alloy, by approximating datatypes
by finite subterm-closed substructures. The approach produces sound (fragments
of) models for formulas in the \relax{existential--bounded-universal} fragment
(i.e., formulas whose prenex normal forms contain no unbounded universal
quantifiers ranging over datatypes). This idiom was further developed by Dunets
et al.\ \cite{dunets-et-al-2010}, who showed how to handle primitive recursion.
Their definedness guards play a similar role to the existential constraints
generated by the $\conv$ translations.

The higher-order model finder Nitpick for the Isabelle/HOL proof assistant uses
another variant of Kuncak and Jackson's approach inside a Kleene-style
three-valued logic. It was also the first tool of its kind to corecursion and
codatatypes \cite{blanchette-2010-tap}. The three-valued logic approach
enriches each type with an unknown value ($\bot$), which is propagated by
function application. This approach works reasonably well in Nitpick, because
it builds on Kodkod and its relational logic, but our initial experiments with
\cvc convinced us that it is more efficient to avoid unknown values by adding
constraints.

The Leon system \cite{leon} implements a procedure that can both produces
proofs and counterexamples for properties of terminating functions written in a
subset of Scala. Leon is based on an SMT solver. It avoids quantifiers
altogether by unfolding recursive definitions up to a certain depth, which is
increased iteratively as necessary.
\jb{Andy: Can you add some more here? You're the Leon pro!}

%Unlike an SMT solver, Nitpick cannot rely on a built-in notion of
%a datatype; it axiomatizes finite subterm-closed substructures.

%\begin{verbatim}
%Nitpick
%  * nice: three-valued logic (e.g. local overflow)
%    * solving a different problem:
%      * no built-in notion, what's approximated is an idealized notion, no
%        injections $\concret$
%  * not so nice:
%    one abstract domain per type, not per function argument
%  * guards + one-domain per arg are a good combi, guards + one-single-domain is bad because of
%    odd schemes, give unsat example
%    [* show that it works also for example with weird recursion schemes]
%
%  * other features:
%    * inductive / coinductive predicates
%    * quotient types
%
%  * abstract interpretation
%
%Leon
%\end{verbatim}

@@

We presented a translation scheme that makes more SMT problems amenable to
finite model findings in SMT. In conjunction with the finite model finding
techniques introduced by Reynolds et al.\
\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013}, it makes it possible to
find models of quantified formulas over infinite types, such as integers
and recursive datatypes.

In future work, it would be interesting to evaluate the approach against other
model finders, notably Leon and Nitpick, and enrich the benchmark suite with
problems exercising \cvc's support for codatatypes
\cite{reynolds-blanchette-2015-codata}. We also plan to integrate \cvc as a
counterexample generator in proof assistants.
